---
title: ''
author: "Grace Lewin"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm Up

```{r}
us_rent_income_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)
```


## Practice some tidyverse functions for data wrangling

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping things together in sequence

### `dplyr::filter()` to keep or exclude rows based on our conditions

1. Exact match (single condition)

Only keep observations (rows) where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```

2. "and" statement within filter

Only keep observations if the penguin species is Chinstrap AND the year is 2008

```{r}
# This first one is most used
ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == 2008)

ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)

ex_2 <- penguins %>% 
  filter(species == "Chinstrap") %>% 
  filter(year == 2008)
```

3. Exclude Gentoos

Keep observataions is the penguin species is NOT "Gentoo"

```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Exercise: Create a subset from `penguins` that only contains observations for Adelie pengins with a flipper length greater than or equal to 185mm.

```{r}
adelie_greater_185 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```

4. The %in% operator

Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row. 

We want to keep observations where the island is Dream or Torgersen.

```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
```

Don't use `==` in this case because it would be looking for where Dream and Torgersen happen subsequently. One after the other.
%in% is looking for *anywhere* where Dream returns TRUE or Torgersen returns TRUE.

### select() to select or exclude COLUMNS

5. Keep a single column by name

Only want to keep the species column

```{r}
ex_5 <- penguins %>%  select(species)

# Reminder: rewrite this without the pipe operator

ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year.

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

7. Choose or exclude columns based on position

Exclude column 4
```{r}
ex_7 <- penguins %>% 
  select(-4)
```


8. A piped sequence

Write a piped sequence where you first: only keep columns for species, body_mass_g, and island, and second: filter to only include observations if body_mass_g is greater than 3600
```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

*Plan on not using `select()` too much because it's not necessary to exclude columns to do analysis on other columns*

### mutate() to add or update existing columns, while deeping the ones already there.

9. Add a column that has body mass converted to kg

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)
```

10. Mutate to update an existing column

```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
```


11. Update the class of the bill depth column to character

a common use of mutate is to update class of columns.
```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm)
```

### case_when(): your favorite if-else

12. Create a new column that contains "small" if penguin flipper length is less than 185 m, "medium" otherwise.

```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: Stariting from penguins, create a new column called "bird_bill_bin" that contains :
- If bill length < 38mm: "short bill"
- If bill length >= 38mm & < 43mm: "medium bill"
- If bill length >= 43: "long bill"

```{r}
practice_1 <- penguins %>% 
  mutate(bird_bill_bins = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))

practice_1 <- penguins %>% 
  mutate(bird_bill_bins = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    TRUE ~ "long bill" # if anything else, populates with "long bill" (but also nas if not removed first). This statement has to come last. Otherwise it will give "long bill" for everything if it is put first.
  ))

# when using case_when, go more specific to least specific.
```


### group_by() %>% summarize() to find summary values by group, and return them in a nice summary table.

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```

13. group_by() to say there are groups within the species column

```{r}
ex_13 <- penguins %>% 
  group_by(species, island, year, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

penguins %>% 
  group_by(species) %>% 
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
## adds a column to the end of penguins. This is DIFFERENT to what you get with the above code. 
## summarize() is NOT the same as mutate()
```

## Practice

In a single piped sequence:

- Only keep penguins that have a body mass > 3500 g
- Only keep columns for species, flipper_length_mm, and sex
- Create a summary table that contains the mean, sd, max and min value of flipper length by species and sex (as the grouping variables)

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE),
            min_flipper_length = min(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup()
```


